---
title: "Programming assigment 4"
subtitle: "Data science"
author: "Jorge Vargas-Mutizabal"
date: "2025/03/29 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# Libraries
```{r}
library(tidyverse)
library(dplyr)
library(ds4ling)
library("here")
```
---
#Data load
```{r}
data <- read.csv(here("data_raw", "fricatives", "fricatives.csv"))
```
---
##Data tidying: COG and Skewness for S and Sh 
```{r}

data_1 <- data |>
  pivot_longer(cols = c(s_cog, sh_cog),
               names_to = "phoneme", 
               values_to = "cog")|>
pivot_longer(cols = c(s_skewness, sh_skewness),
             names_to = "phoneme_skewness",
             values_to = "skewness")|>
  select("phoneme", "cog", "skewness")|>
  mutate(phoneme = gsub("_cog", "", phoneme))
  
```
---
# Data print
```{r}
print(data_1)
```


---
#Table of descriptive statistics
```{r}

summary_table <- data_1 |>
  group_by(phoneme)|>
  summarise(
    mean_skew = mean(skewness),
    mean_cog = mean (cog),
    mean_skew = median (skewness),
    mead_cog = median (cog)
  )

knitr::kable(summary_table)

  
```
---
#Tidy data
```{r}
write.csv(data_1, "../data_tidy/data_tidy.csv")
```


---
#Boxplot of COG as a function of phoneme
```{r fig.height=5, fig.width=5}
data_1|>
  ggplot() +
  (aes(x = phoneme, y = cog, fill = phoneme)) +
  geom_boxplot() +
  labs(title = "Center of gravitiy",
       x = "Phoneme", y = "Center of gravity")
```

---
## Plot skewness as a function of phoneme
```{r fig.height=5, fig.width=5.5, warning=FALSE}
  ggplot(data_1, 
         aes(x = phoneme, y = skewness, fill = phoneme)) + 
  stat_summary(fun.data = mean_sdl) +
  labs(title = "skewness as a function of phoneme",
       x = "Phoneme", y = "Skewness")

```
---
## COG as a function of skewness for the [s] segments 
```{r fig.height=4, fig.width=7}
data_s <- data_1 |>

  filter(phoneme == "s")  

model <- lm(
  cog ~ skewness,
  data = data_s)

print(model)
```
---

##Scatter plot of COG ~ skweness for [s]
```{r fig.height=3.5, fig.width=5, message=TRUE}
data_s |>
  ggplot() +
  aes(x = cog, y = skewness) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Gravity as a function of skwness",
       x = "COG", y = "Skewness")
```
---
##Table model: Coefficients
```{r}
knitr::kable(summary(model)$coef)
```

---
##Model diagnostics

```{r fig.height=6.5, fig.width=7}
diagnosis(model)
```
---
#Report
Center of gravity (COG) data of the phoneme [s] was analyzed as a function of skewness using a linear model. The model included COG [s] as the dependent variable and skewness as the independent variable. The results showed a positive correlation between the two variables. The COG increases along the level of skewness in the phoneme s, [B = 995.87, SE = 111.14 , p = 1.31e-13]. The results suggests that COG increases 995.87 per unit point of skewness, and it is statistically significant. 


---
# Assumptions data
```{r}
data_2 <- read_csv("../data_raw/fricatives/assumptions.csv")
```
---
##Plotting data: Surprise!
```{r fig.height=5, fig.width=5}
data_2|>
  
  ggplot() +
  geom_point() +
  aes(x = x, y = y) +
  labs (x = "x", y = "y", tile = "Assumptions")
```
---
# Explanation
The data does not have real data. It makes a cute dinasaur. 

---
# ¡Muchas gracias!
